\documentclass[../main.tex]{subfiles}
\begin{document}
\chapter{Implementation}
%
%
%
\section{Semantic Highlighting}
\subsection{How a Language Server highlights}
\todo[inline]{Research this in more detail}

\subsection{Template Parsing}
First, the lines containing templates are found. These are found by starting after the header \codeword{the templates are:}, continuing until either another header or the end of the document is reached. A corresponding template is constructed from each non-empty line by the template class: sub-strings wrapped in asterisks are taken to contain types, and the relevant \codeword{Type} object is put in place, taken from the type tree.

\subsection{Extracting the terms of a literal}
The fundamental problem here is, given a template \codeword{T} and a literal \codeword{L} that is assumed to match, how to extract the elements of the literal? In short, this is done by leveraging the assumptions that the template and the literal share the same surroundings. Under this assumption, comparing the template's surroundings against the literal yields the literal's terms. The algorithm for this is given below.
\todo[inline]{Write up algorithm here}
This algorithm was constructed using the idea of incremental problem solving in my Algorithsm course.
There is a slight subtlety here: what happens if a surrounding also appears as a term. For example, a scenario where a merchant packages and sends items could be described with the template \codeword{*a merchant* ships *an item*}. If we have a merchant who packages and sends ships, then the corresponding literal would be \codeword{the merchant ships ships}.
\todo[inline]{Describe how this nuance is handled}
\todo[inline]{Talk about how this algorithm works with the edge case of incomplete literals also}

\subsection{Matching a template to a literal}
Once all the template representations have been created, the templates can be used to highlight the terms in each literal. Starting with a given literal, the templates are filtered by whether or not they match the literal. 
For each template, once the literal's elements have been extracted, the elements are checked to see whether they are isomorphic to the template's elements. That is, whether the surroundings match up between them.

\subsection{Finding the best match}
Old design: takes first template that matches. Problematic, especially with `default templates' e.g. ``*a thing* is *a thing*''
\\
New design with ranking \textit{partially written} templates according to how well they match
\\
\\
Initially, it was assumed that a literal can only match one template. As the editor was being developed, I soon saw how this was often false. This was most clearly visible when ``default'' templates were implemented -- general templates, such as \codeword{*a thing* is *a thing*} that were implicitly present in every Logical English document. With the above example, if a user then supplies a template \codeword{*a beneficiary* is included in *a will*}, then any literal that matches the latter template -- for example, \codeword{jane is a beneficiary of the will} -- will also match the former template. However, using the former template to extract terms will be incorrect, such as highlighting \codeword{a beneficiary of the will} as a single term.
\\ 
\\
This motivated a `match score' between a literal and a template. The higher the score, the better the template matches the literal, and the more certain we can be that it should be used to extract the terms.
\todo[inline]{Write up scoring algorithm}
\todo[inline]{Con: this is not normalised -- how are comparisons then meaningful?}
%
%
%
\section{Completion}
\subsection{How a language server completes code}
\todo[inline]{Research this in more detail}

\subsection{Completing the remainder of a literal}
To offer completion for a literal, first its corresponding templates are found. This is not as straightforward as searching for which templates match the literal, because the literal will be incomplete, and so will not match any template. Instead, the templates are ranked by their match score against the literal, and the top three are taken.
\\
\\
There is already some nuance here. Some matches will be ``obviously'' irrelevant, such as (the dreaded) \codeword{*a thing* is *a thing*} against \codeword{the person is a beneficiary of }. These cannot be ruled out algorithmically, however, since they do match. They will, however, be out-ranked by templates with better-matching surroundings, and will appear lower in the list.
Finding best three templates according to match score. Filling in templates with terms that the literal has so far.
\todo[inline]{Justify why at most three literals are suggested. Research user design.}
%
%
%
\section{Error diagnosis}
\subsection{How a language server diagnoses errors}
\todo[inline]{Research this in more detail}

\subsection{Diagnosing literals that have no matching template}
Given the above discussion it is fairly straightforward to find literals that do not match any template. The nuance is in using this as the deciding factor actually being too eager. Specifically, incomplete literals (e.g. literals that are being typed) will, in general, not match a template. This means that every literal that is being typed will be diagnosed as incorrect until it is finished.
\todo[inline]{Talk about how this is impossible to fix: on diagnosis time, all that is given is the text document, not the cursor position.}

\subsection{Diagnosing clauses that have misaligned connectives}
This is another problem that sounds simple in theory, once the above framework was developed, but turned out to have nuances in practice. Each clause is split into its lines, and the lines containing the keywords \codeword{and} and \codeword{or}, which have equal precedence, have their indentation compared. If two literals with different connectives have the same whitespace, then the precedence of the connectives is ambiguous.
\\
\\
The nuance is with the term `same indentation'. There are two equally common ways to indent lines: tabs and spaces. There is no standard way to display a tab in terms of spaces: any range from two to eight spaces is common. Thus, if one person indents literals using tabs, and the other using spaces, then it is up to interpretation as to whether the indent is correct or not.
\todo[inline]{Cite sources on this. Talk about how this is a problem in Python, and that there is no solution. Talk about how it should therefore be an error to mix spaces and tabs, but I do not want to check for this.}
Issue: spaces versus tabs

\subsection{Diagnosing type mismatches}

\subsubsection{Feature Overview}
Although not part of the editor's requirements, the Logical English development team wanted a way to introduce experimental support for typed terms. The argument labels in templates would give the type of their argument. This feature would be used for finding type errors in multiple uses of the same term across a clause. Since this feature is experimental, it was not supposed to clash with any existing features or hinder the experience of a user who did not want this.
\\
\\
The way I chose to implement this feature was to have type checking turned off by default, only being turned on with an explicit \codeword{type checking: on} comment. The type hierarchy was also designed to be as minimal as possible. 
\todo[inline]{Move this discussion to Requirements section.}

\subsubsection{Initial design: flat type hierarchy}
When experimenting with implementing this feature, I first implemented type checking before implemnting a type hierarchy. In each clause, the literals were extracted, and, using the document's templates, the terms were extracted from each literal. These terms were typed, but there was no notion of sub-type or super-type. If two terms were found that had the same name but different types, a type mismatch error message was generated.
\\ 
\\
This design was a lot more inconvenient to use than I expected, with many more surprising error messages being generated. This was mainly because of the default templates, and for two reasons. Firsly, since the default templates are very broad, they use placeholder type names, such as \codeword{A}, \codeword{B}, \codeword{C}, \codeword{thing}, et cetera. Ths caused problems for two reasons. Firstly, the type names clashed with the more specific type names found in real-world Logical English examples, if they were used together. Secondly, the type names often clashed amoungst themselves: take, for example, the template \codeword{*an A* appended to *a B* gives *a C*}, and the literal \codeword{[] appended to the list gives the list}. Here \codeword{the list} is both of type \codeword{B} and \codeword{C}.
\todo[inline]{Cite the default templates, or at least, talk about them in the LE specification.}
The solution to the first problem could only be resolved through a type hierarchy, making sure that the types used in the default templates would be super-types of whatever is used in more specific templates. However, the solution to the second problem was immediately resolvable: rename the types. All general type names such as \codeword{A}, \codeword{B}, \codeword{C} et cetera were renamed to \codeword{thing}.

\subsubsection{Diagnosing with a type hierarchy}
First, the type tree is read from the document under the header \codeword{the type hierarchy is:}. Parsing a text version of a tree is a well-known algorithmic problem.
\todo[inline]{Cite the solution.}
Now instead of simply checking whether two type names are equal, the two types are checked to see if one is a subtype of the other. Since types are rerpesented in a tree, this is a simple tree search problem to see whether either type node has a child node with the same name as the other type.
\todo[inline]{Cite the solution.}
%
%
%
\section{Quick Fixes}
\subsection{How a language server provides quick fixes}
\todo[inline]{Research this in more detail}

\subsection{Fixing a lack of template for literals}
\subsubsection{Template Generation using Least General Geneneralisation}
The first candidate template is generated according to the principle of least general generalisation. This is a principle from Logic-Based Learning which states how to generate a predicate that matches a set of given instantiations of the predicate, that is no more general than is necessary to match all of them.
\todo[inline]{Research and describe this more.}
The way this is done in the editor is as follows. Each literal is split into a list of space-separated words. It is assumed that there is a common template which matches all the literals. In the initial design, the predicate words were identified as the intersection of all the words. This criteria is slightly refined, and will be discussed later.
\\ 
\\
Having found all of the predicate words, the terms of the first literal are identified as the words that are not predicate words. Knowing both the predicate words and the terms of the first literal, I then generate a template that matches this. If this same template matches all the other literals, then it is returned.
\todo[inline]{Write up the algorithm.}

\subsubsection{Template Refinement from the re-use of terms}
It was often the case that the least general generalisation was not enough to generate an accurate template. Either there were too few example literals, or the examples did not vary every term. In trying to fix this problem, I noticed that I was not exploiting the context(i.e. the clause) in which the literal was written. If a literal borrows a term from another literal, then we know about that term, and can generalise it into a variable.
\\
\\
This was done by giving the \codeword{Template} class a method to generate a more general template by generalising a given term into a variable. Applying this method successively to all the surrounding terms that feature in each template-less literal gave much more accurate templates. This also allowed a single literal to be generalised into a template -- a feature that is impossible with least general generalisation. This also told us the type to place inside the template, since the type of each term is known.

\end{document}