\documentclass[../main.tex]{subfiles}

\begin{document}
\chapter{Introduction}
\section{Abstract}
Language Extensions for code editors are a crucial tool in writing code quickly and without errors. In this project, I create a language extension for the logical, declarative programming language Logical English. The language extension highlights the syntactic and semantic features of Logical English, predicts the completion of atomic formulas, identifies errors and generates ``boilerplate'' code to fix them. The language editor also extends the language of Logical English through type checking the use of terms by introducing a hierarchical type system. The extension is based on a language server that uses the Language Server Protocol. It is evaluated when connected to a Visual Studio Code front-end.

\section{Purpose of this Document}
This document servers two purposes. The first is to introduce Logical English and the editor to new users, setting the scene for the later analysis. This is done in Part 1, which sets the scene for the later sections. The second purpose is to describe the requirements, research, design and implementation of the editor, evaluate the finished product, and lay out how this editor is connected to the wider body of surrounding literature. This is done in Part 2 of the document.

\section{Motivation: why a new editor?}
Logical English is a relatively new programming language, first introduced under that name in late 2020 \cite{logical_english}. Although Logical English has an online editor hosted on the SWISH platform \cite{swish_editor}, the editor lacks features to edit Logical English that are common to most programming language editors. The SWISH editor is primarily a Prolog editor, so any Logical English code has to be written in a string that is an argument to a Prolog function. Since Logical English code is written in a Prolog string, the Logical English content cannot be treated by the editor as a standalone program. This means that it can receive no syntax highlighting, error detection, or code completion features beyond that of a Prolog string. Since these features are essential for productivity, a new editor was needed that was custom-built for writing Logical English.
\todo[inline]{Find a source about the productivity of IDEs vs a plaintext editor}

\section{Editor Overview}
The editor documented in this report is a language extension for the popular Integrated Development Environment (IDE) `Visual Studio Code'. The language extension consists of three components: a language client, a syntax highlighter, and a language server. 

\subsection{Language Client}
The language client is built for Visual Studio Code. The language client delivers the editor's features to the user's Visual Studio Code window.

\subsection{Syntax Highlighter}
The syntax highlighter is a document that contains the grammar of Logical English. The language client uses the syntax highlighter to highlight aspects of Logical English in the user's document. The Syntax Highlighter is written according to the TextMate grammar specification, allowing it to be used by language clients for IDEs other than Visual Studio Code.

\subsection{Language Server}
The language server calculates the main features of the editor. When the user writes in the document, the language client communicates the change to the language server. The language client may request features such as:
\begin{itemize}
    \item whether there are any errors in the document to diagnose
    \item whether what the user is currently writing can be auto-completed
    \item if the user is hovering over an error message, whether a `quick fix' can be suggested
\end{itemize}
The language server calculates these features and communicates them to the language client. The language client ensures that these features are then displayed to the user. 
\\
\\
The language server is written according to the Language Server Protocol. This allows the language server to be used by language clients for many IDEs other than Visual Studio Code.
\end{document}